---
title: "Getting Started with PREVAIL"
author: "The PREVAIL Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
## Load package when installed, or via devtools when developing
if (requireNamespace("PREVAIL", quietly = TRUE)) {
  library(PREVAIL)
} else if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all()

  
}

# library(tidyverse)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
options(scipen = 999)
```

# Introduction

Welcome to the `PREVAIL` package — **PRojection of Epidemics and VAccination Impact under Lapses in coverage** — a flexible framework for age-, risk-, and vaccination-stratified infectious disease modeling, tailored for crisis settings. This vignette walks through a basic end-to-end simulation using default package functions to load data, run a stochastic transmission model, and visualize results.

In this vignette we will be using the default data included in the package. If you are interested in learning how to use your own data on [demography](), [vaccination](), or [infection dynamics]() please see the linked vignettes.

The below example will run through a measles transmission model in Palestine ("PSE") using these prebuilt internal datasets and the transmission model implemented in [Odin](https://mrc-ide.github.io/odin2/).

# 1. Install and load Required Packages

First we need to install the **PREVAIL** package, which we install from github using the function `install_github()` from the **devtools** package.

```{r, eval = FALSE}
devtools::install_github("arranhamlet/PREVAIL")
```

Once installed, we can load in the required packages for this vignette. In addition to **PREVAIL**, we will also be using **tidyverse**. To load these, we will be using the **pacman** package, which will also install these packages if missing.

```{r, eval = FALSE}
#If pacman is not already present, install
if(!require("pacman")) install.packages("pacman")

#Install missing packages, and load 
pacman::p_load(
  PREVAIL,
  tidyverse
)

```

# 2. Load and Process Parameters

To run our model, we need to first prepare our parameters. To do this, we use the function `data_load_process_wrapper()`. This function loads the data included in the package on demography, vaccination coverage, and disease dynamics. In order to run the model, we just need to specify a few arguments.

**iso** - The 3 letter country code that specifies which UN recognised country or territory we are interested in. This will determine our demography, vaccination coverage and prior disease exposure. For a full list please see `data(PREVAIL_locations)`. In this example we will use "PSE" for Palestine.

**R0** - The basic reproductive number. This determines how many secondary infections we would expect to see from an initial infection. The higher the number, the more infectious the disease is.

**year_start** - The first year of data to input into the model. This determines the data to load into the model, and allows us to replicate historical patterns in demography, vaccination and infection. The range is from 1950 to 2023, and the default of "" uses the entire dataset.

**year_end** - The final year of data to input into the model. This determines the data to load into the model, and allows us to replicate historical patterns in demography, vaccination and infection. The range is from 1950 to 2023, and the default of "" uses the entire dataset. This must be equal to or larger than `year_start = .`

**WHO_seed_switch** - If FALSE, the model allows for dynamic disease transmission throughout the entire period. If TRUE, the model allows for dynamic disease transmission prior to 1980 (when WHO cases are not available), but from 1980 onwards, it matches annual cases to the number reported to the WHO. This allows the user to replicate reported cases.

**aggregate_age** - If TRUE, the model aggregates the baseline single year age groups of 0-100, to the values specified in `new_age_breaks = `.

**new_age_breaks** - The age breaks to aggregate to if `aggregate_age = TRUE`.


```{r, eval = FALSE}
params <- data_load_process_wrapper(
  # Specify the country or territory using its ISO3 code.
  # This determines demography, vaccination coverage, and prior infection.
  iso = "PSE",  # Palestine

  # The disease of interest; used to match internal schedules and parameters.
  disease = "measles",

  # Basic reproductive number; governs overall infectiousness.
  R0 = 15,

  # Start year of historical data input (1950–2023).
  # Set to "" to use the full dataset range.
  year_start = "",

  # End year of historical data input (1950–2023).
  # Must be equal to or after year_start.
  year_end = "",

  # If TRUE, uses WHO-reported annual case data from 1980 onward to seed the model.
  # If FALSE, allows fully dynamic transmission throughout.
  WHO_seed_switch = TRUE,

  # Whether to aggregate single-year age groups (0–100) into larger age bands.
  aggregate_age = TRUE,

  # The age group cut points to use if aggregating age groups.
  new_age_breaks = c(0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, Inf)
)

```

The output of this function is a list of the parameters required to run the mathematical model of disease transmission. 

# 3. Run the Model

One we have generated our parameters, we can use the function `run_model_unpack_results()` to run the model using our previously generated parameters.

This function takes two arguments, the parameters we have previously generated `params`, and the `no_runs` which specifies the number of stochastic runs we want the model to run. Running the model multiple times helps account for the inherent randomness in disease transmission. However, increasing the number of runs also increases computation time. For large populations (e.g., over one million), results tend to be stable across runs, so a smaller number of simulations is often sufficient. In contrast, smaller populations (e.g., a few thousand) may require more runs to capture variability in outcomes.

Here we will look at a single run of the model.

```{r, eval = FALSE}
model_run <- run_model_unpack_results(
  params = params,
  no_runs = 1
)
```

# 4. Check our model outputs

Once run, we can quickly look to see if the model has accurately captured the total population, and the end age breakdowns over time. This is a useful check to make sure we are going to be accurately capturing the effects of historical vaccination and infection on our current population.

To do this we use the function `demographic_check()` which will produce two plots. 

1) The total population over time from our model and a reference population.
2) The number of individuals by age from our model and a reference population for the final year of the simulation.

This function takes three arguments:

**reference_population** - The population we want to compare our model results to. This is provided by our `data_load_process_wrapper()` function and can be extracted in our vignette through `params$population`. This must match the age compartments found in the model.

**model_run** - The output of `run_model_unpack_results()`.

**age_breaks** - A numeric vector of age group breakpoints (e.g., `c(0, 5, 10, ..., 80, Inf)`). Found in `params$input_data$age_breaks`.

```{r, eval = FALSE}
#Check demographics
demographic_check <- demographic_check(
  reference_population = params$population * 1000,
  model_run = model_run,
  age_breaks = params$input_data$age_breaks
)
```


## Visualising Model Outputs

Once the model is run, we can quickly visualise the outputs using the function `summary_plots()`, which produces four plots and a data frame summarising susceptibility profiles over time and age.

```{r, eval = FALSE}
plots <- summary_plots(
  model_run = model_run,
  params = params
)
```

The function returns a named list containing:

- `select_state_plot`: A faceted time-series plot showing the median and 95% uncertainty intervals for:
  - Susceptible
  - Exposed
  - Infectious
  - Infectious, severe
  - Recovered
  - Recovered with complications
  - Total population
  - Cases

- `susceptibility_plot`: A bar chart showing the susceptibility profile of the population at the most recent year, broken down by age group.

- `vaccination_coverage_plot`: A line plot showing the proportion of the population with any vaccine-derived protection (including mixed) over time.

- `cumulative_case_plot`: A cumulative incidence plot with uncertainty bands.

- `susceptibility_data`: A weekly `data.table` in long format that shows the evolving proportion of the population in each susceptibility class.

### Susceptibility Categories

The population is assigned one of four labels based on vaccination and infection history:

1. **Susceptible** – Never vaccinated and never infected.
2. **Vaccine derived protection** – Vaccinated, but no infection history.
3. **Infection derived protection** – Infected, but never vaccinated.
4. **Vaccine and infection derived protection** – Both vaccinated and previously infected.

These labels are calculated across time, age, vaccination status, and risk group, and aggregated into interpretable summaries annually and weekly.


# 5. Calculate current susceptibility and prepare for onward projections.

Once we are satisfied with our model outputs, we can calculate the susceptibility of the population at the final timepoint. This will be used to establish a baseline population-level susceptibility for our onward projections. This is done using `current_susceptibility()`, which uses the output of `run_model_unpack_results()`.

```{r, eval = FALSE}
#Process current susceptibility for onward runs
current_susceptibility <- calculate_current_susceptibility(
  model_run = model_run
)
```

Next, we want to set up our onward scenario projections. In this example we will be looking forward 10 years, with vaccination steadily decreasing, cases being introduced in year 7, and R0 remaining constant. To summarise our onwards scenario, we need to create a data.frame with four columns.

1) **year** - The year of simulation.
2) **relative_coverage** - The relative vaccination coverage compared to the final entry of `params$vaccination_coverage`. A value of 0.5 would indicate vaccination coverage is half of previously seen, and a value of 2.0 would indicate vaccination coverage is twice as high.
3) **introduced_cases** - The number of cases to be introduced in that year. Introduced at the first timestep of the year and allocated to the unvaccinated compartment. Age is randomly selected based on a population weighting.
4) **R0** - The absolute value of R0.

Here we will create an example of when vaccination activities reduce over time, falling to 50%, then 25% and then to 0% vaccination coverage relative to our initial timepoint.

```{r, eval = T}
vaccination_reduction <- data.frame(
  year = 0:9,
  relative_coverage = c(0.5, 0.5, 0.5, 0.25, 0.25, 0.25, 0, 0, 0, 0),
  introduced_cases = c(0, 0, 0, 0, 0, 0, 0, 10, 0, 0),
  R0 = 15
)

vaccination_reduction
```

We are also going to run a baseline example, where vaccination coverage is maintained (`relative_coverage = 1`) for comparison.

```{r, eval = F}
vaccination_maintenance <- data.frame(
  year = 0:9,
  relative_coverage = 1,
  introduced_cases = c(0, 0, 0, 0, 0, 0, 0, 10, 0, 0),
  R0 = 15
)
```

To model this forward projection we use the function `prepare_future_data()` which will take our original parameters generated with data_load_process_wrapper() `params`, the current population susceptibility, generated with `current_susceptibility()`, and our data.frame objects of future events.

```{r, eval = F}

reduction_params <- prepare_future_data(
  params = params,
  current_susceptibility = current_susceptibility,
  future_events = vaccination_reduction
)

maintenance_params <- prepare_future_data(
  params = params,
  current_susceptibility = current_susceptibility,
  future_events = vaccination_maintenance
)

```

# 6. Onward projection and analysis

Once parameterised for our onward projections, we can use the previously described `run_model_unpack_results()` and `summary_plots()` to run our model and produce summary plots using our new parameters.

```{r, eval = F}
run_reduction <- run_model_unpack_results(
  reduction_params,
  simulation_length = 10 * 365,
  no_runs  = 5
)

#Future results
run_maintenance <- run_model_unpack_results(
  maintenance_params,
  simulation_length = 10 * 365,
  no_runs  = 5
)
```

From here we can generate plots using `summary_plots()` which also contain information that we will use in a function `scenario_compare()` to compare the model runs. This function takes two arguments, `scenario_1` and `scenario_2` which are the outputs of `summary_plots()`.

```{r, eval = F}

#Generate plots
reduction_plots <- summary_plots(
  model_run = run_reduction,
  params = reduction_params
)

maintenance_plots <- summary_plots(
  model_run = run_maintenance,
  params = maintenance_params
)

```

## Comparing Transmission Scenarios

To compare outputs across different intervention or transmission settings, we use the `scenario_compare()` function. This provides a unified summary of model dynamics and population susceptibility across two model scenarios.

```{r, eval = FALSE}
comparison <- scenario_compare(
  scenario_1 = scenario_baseline,
  scenario_2 = scenario_intervention
)
```

### Function Outputs

This function returns a named list with the following components:

- `summary_plot`: A composite plot combining:
  - A **weekly time-series** of median cases for both scenarios with uncertainty ribbons.
  - A **cumulative case trajectory** plot showing the build-up of cases over time.
  - A **susceptibility trajectory** over time, highlighting changes in the proportion of the population classified as susceptible.
  - A **summary table** comparing scenario statistics (peak, total cases, and susceptibility increase).

- `summary_table`: A formatted `gt` table that includes:
  - **Peak median cases** and the **time of peak**.
  - **Total cases** with uncertainty bounds.
  - The **percentage increase in susceptible population** at the timepoint of maximum difference between scenarios.

- `individual_plots`: A named list of individual `ggplot2` objects for further use, including:
  - `case_plot`: Weekly incidence of median cases.
  - `cumulative_plot`: Cumulative cases with uncertainty intervals.
  - `susceptibility_plot`: Trajectory of susceptible population proportion over time.

### Interpreting the Outputs

This function is designed for side-by-side comparison of different scenario runs from the same model structure.

The **case plots** illustrate both the scale and timing of outbreaks, allowing visual inspection of:
- **Delays** in peak infection
- **Reduction** in overall burden
- **Shifts** in outbreak shape due to interventions

The **susceptibility plot** shows the dynamics of the proportion of the population still at risk over time. This is particularly useful for evaluating how vaccination or prior immunity influence long-term vulnerability.

The **summary table** allows easy extraction of numerical differences, including:
- Differences in total and peak cases.
- Time shifts in the epidemic peak.
- Increase or reduction in susceptibility over time, useful for understanding potential rebound risks.

> **Note:** Both scenarios must include the same structure of outputs — including `select_state_plot`, `susceptibility_data`, and `cumulative_case_plot`. Each scenario should be the result of a full call to `summary_plots()` or similar.

